---
title: "Chapter 3 - Sampling the Imaginary"
format: html
---

```{r}
library(rethinking)
library(posterior)
library(cmdstanr)
library(tidyverse)

theme_set(theme_minimal())
```


```{r}
Pr_Positive_Vampire <- 0.95
Pr_Positive_Mortal <- 0.01
Pr_Vampire <- 0.001
Pr_Positive <- Pr_Positive_Vampire * Pr_Vampire + Pr_Positive_Mortal * (1 - Pr_Vampire)
Pr_Vampire_Positive <- Pr_Positive_Vampire * Pr_Vampire / Pr_Positive

Pr_Vampire_Positive
```

## 3.1 Sampling from a grid-approximate posterior

```{r}
p_grid <- seq(from = 0, to = 1, length.out = 1000)
prior <- rep(1, 1000)
likelihood <- dbinom(x = 6, size = 9, prob = p_grid)

unstd_posterior <- likelihood * prior
posterior <- unstd_posterior / sum(unstd_posterior)
```

```{r}
samples <- sample(x = p_grid, size = 10000, replace = TRUE, prob = posterior)

samples_data <- tibble(
    sample_number = 1:length(samples),
    samples = samples
)

samples_data |>
    ggplot(aes(x = sample_number, y = samples)) +
    geom_point(alpha = 0.4)

samples_data |>
    ggplot(aes(x = samples)) +
    geom_density()
```

## 3.2 Sampling to summarize

### 3.2.1 Intervals of defined boundaries
```{r}
sum(posterior[p_grid < 0.5])
```

```{r}
sum(samples < 0.5) / 1e4
```

```{r}
sum(samples > 0.5 & samples < 0.75) / 1e4
```

```{r}
quantile(samples, 0.8)
```

## 3.5 Practice

```{r}
p_grid <- seq(from = 0, to = 1, length.out = 1000)
prior <- rep(1, 1000)
likelihood <- dbinom(6, size = 9, prob = p_grid)
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)

set.seed(100)
samples <- sample(x = p_grid, size = 1e4, replace = TRUE, prob = posterior)
```

### 3E1
How much posterior probability lies below $p = 0.2$?

4e-04.
```{r}
mean(samples < 0.2)
```

### 3E2
How much posterior probability lies above $p = 0.8$?

11.16%.
```{r}
mean(samples > 0.8)
```

### 3E3
How much posterior probability lies between $p = 0.2$ and $p = 0.8$?

88.8%.

```{r}
mean(samples >= 0.2 & samples <= 0.8)
```

### 3E4
20% of the posterior probability lies below which value of $p$?

$p = 0.52$.
```{r}
quantile(samples, 0.2)
```

### 3E5
20% of the posterior probability lies above which value of $p$?

$p = 0.76$.
```{r}
quantile(samples, 0.8)
```

### 3E6
Which values of $p$ contain the narrowest interval equal to 66% of the posterior probability?

$p = 0.51$ and $p = 0.77$.
```{r}
bayestestR::hdi(samples, 0.66)
```

### 3E7 
Which values of $p$ contain 66% of the posterior probability, assuming equal posterior probability both below and above the interval?

$p = 0.50$ and $p = 0.77$.
```{r}
bayestestR::eti(samples, 0.66)
```

### 3M1
Suppose the globe tossing data had turned out to be 8 water in 15 tosses. Construct the posterior distribution, using grid approximation. Use the same flat prior as before.

```{r}
# Making some functions to ease up on the amount of code I have to right for the exercises
uniform_prior <- \(p_grid) rep(1, length(p_grid))
truncated_prior <- \(p_grid) if_else(p_grid < 0.5, 0, 1)

grid_approximation <- function(grid_size, prior_func = NULL, w, n) {
    p_grid <- seq(from = 0, to = 1, length.out = grid_size)
    prior <- prior_func(p_grid)

    likelihood <- dbinom(w, n, p_grid)
    posterior <- likelihood * prior
    posterior <- posterior / sum(posterior)

    posterior
}
```

```{r}
posterior <- grid_approximation(1000, uniform_prior, 8, 15)
samples <- sample(x = p_grid, size = 1e4, replace = TRUE, prob = posterior)
```

```{r}
data.frame(samples) |>
    ggplot(aes(x = samples)) +
    geom_density()
```

```r
# This also does the same - using logic from Chapter 2
data.frame(
    p_grid = p_grid,
    posterior = posterior
) |>
    ggplot(aes(x = p_grid, y = posterior)) +
    geom_line()
```

### 3M2
Draw 10,00 samples from the grid approximation above. Then use the samples to calculate the 90% HDPI for $p$. 

The 90% HDI for $p$ is $[0.33, 0.72]$.
```{r}
bayestestR::hdi(samples, 0.9)
```

### 3M3
Construct a posterior predictive check for this model and data. This mean simulating the distribution of samples, averaging over the uncertainty in $p$. What is the probability of observing 8 water in 15 tosses?

The probability of observing 8 water in 15 tosses is 0.1444.
```{r}
posterior_pred_dist <- rbinom(n = 1e4, size = 15, prob = samples)
mean(posterior_pred_dist == 8)
```

### 3M4
Using the posterior distribution constructed from the new (8/15) data, now calculate the probability of observing 6 water in 9 tosses.

The probability of observing 6 water in 9 tosses is 0.197.
```{r}
likelihood <- dbinom(6, size = 9, prob = posterior)
new_posterior <- likelihood * posterior
new_posterior <- new_posterior / sum(new_posterior)

samples <- sample(x = p_grid, size = 1e4, replace = TRUE, prob = new_posterior)
posterior_pred_dist <- rbinom(n = 1e4, size = 9, prob = samples)
mean(posterior_pred_dist == 6)
```

### 3M5
Start over at **3M1**, but now use a prior that is zero below $p = 0.5$ and constant above $p = 0.5$. This corresponds to prior information that a majority of the Earth's surface is water. Repeat each problem above and compare the inferences. What difference does the better prior make? If it helps, compare inferences (using both priors) to the true value $p = 0.7$.

### 3M1 Take 2
```{r}
posterior <- grid_approximation(1000, truncated_prior, 8, 15)
samples <- sample(x = p_grid, size = 1e4, replace = TRUE, prob = posterior)
```

```{r}
data.frame(samples) |>
    ggplot(aes(x = samples)) +
    geom_density()
```


### 3M2 Take 2
```{r}
bayestestR::hdi(samples, 0.9)
```

### 3M3 Take 2
```{r}
posterior_pred_dist <- rbinom(n = 1e4, size = 15, prob = samples)
mean(posterior_pred_dist == 8)
```

### 3M4 Take 2

```{r}
likelihood <- dbinom(6, size = 9, prob = posterior)
new_posterior <- likelihood * posterior
new_posterior <- new_posterior / sum(new_posterior)

samples <- sample(x = p_grid, size = 1e4, replace = TRUE, prob = new_posterior)
posterior_pred_dist <- rbinom(n = 1e4, size = 9, prob = samples)
mean(posterior_pred_dist == 6)
```


```{r}

```