---
title: "Chapter 2: Small Worlds and Large Worlds"
format: html
---

```{r}
library(rethinking)
library(tidyverse)
library(cmdstanr)
library(posterior)

theme_set(theme_minimal())
```


## 2.1: The garden of forking data

### 2.1.1 Counting probailities

### 2.1.2 Combining other information

### 2.1.3 From counts to probability

```{r}
ways <- c(0, 3, 8, 9, 0)
ways / sum(ways)
```


## 2.2 Building a model

### 2.2.1 A data story

### 2.2.2 Bayesian updating

### 2.2.3 Evaluate

## 2.3 Components of the model

### 2.3.1 Variables

### 2.3.2 Definitions

#### 2.3.2.1 Observed variables

```{r}
dbinom(6, 9, 0.5)
```

#### 2.3.2.2 Unobserved variables

### 2.3.3 A model is born

## 2.4 Making the model go

### 2.4.1 Bayes' theorem

### 2.4.2 Motors

### 2.4.3 Grid approximation

```{r}
uniform_prior <- \(p_grid) rep(1, length(p_grid))
truncated_prior <- \(p_grid) if_else(p_grid < 0.5, 0, 1)
double_exp_prior <- \(p_grid) exp(-5 * abs(p_grid - 0.5))
```

```{r}
grid_approximation <- function(prior_func, grid_size, successes, trials) {
    p_grid <- seq(0, 1, length.out = grid_size)
    prior <- prior_func(p_grid) 
    likelihood <- dbinom(successes, trials, p_grid)

    unstd_posterior <- likelihood * prior
    posterior <- unstd_posterior / sum(unstd_posterior)

    list(p_grid = p_grid, posterior = posterior)
}
```

```{r}
p_grid <- seq(0, 1, length.out = 20)
prior <- rep(1, 20)
likelihood <- dbinom(6, 9, p_grid)
unstd_posterior <- likelihood * prior
posterior <- unstd_posterior / sum(unstd_posterior)

posterior
```

```{r}
grids <- c(5, 20)
uniform_approximations <- map(.x = grids, 
                      .f = \(grid) grid_approximation(uniform_prior, grid, successes = 6, trials = 9) |> 
                                   as_tibble() |> mutate(points = grid)) |>
                  bind_rows()

uniform_approximations |>
    ggplot(aes(x = p_grid, y = posterior)) +
    geom_line() +
    facet_wrap(~points, scales = "free_y") +
    labs(title = "Uniform prior")
```


```{r}
truncated_approximations <- map(
    .x = grids,
    .f = \(grid) grid_approximation(truncated_prior, grid, successes = 6, trials = 9) |>
        as_tibble() |>
        mutate(points = grid)
) |>
    bind_rows()

truncated_approximations |>
    ggplot(aes(x = p_grid, y = posterior)) +
    geom_line() +
    facet_wrap(~points, scales = "free_y") +
    labs(title = "Truncated prior")
```

```{r}
double_exp_approximations <- map(
    .x = grids,
    .f = \(grid) grid_approximation(double_exp_prior, grid, successes = 6, trials = 9) |>
        as_tibble() |>
        mutate(points = grid)
) |>
    bind_rows()

double_exp_approximations |>
    ggplot(aes(x = p_grid, y = posterior)) +
    geom_line() +
    facet_wrap(~points, scales = "free_y") +
    labs(title = "Double exponential prior")
```


### 2.4.4 Quadratic approximation

```{r}
w <- 6
n <- 9
x <- seq(0, 1, length.out = 100)
true_posterior <- dbeta(x, w + 1, n - w + 1)
```

```{r}
globe_mcmc <- cmdstan_model("Chapter 2 - Small Worlds and Large Worlds/binomial_model.stan")
globe_fit <- globe_mcmc$sample(data = list(N = 9, y = 6), parallel_chains = 4)
```

```{r}
as_draws_df(globe_fit)
```

```{r}
data <- tibble(
    toss = c("W", "L", "W", "W", "W", "L", "W", "L", "W"),
    trial = 1:9,
    n_successes = cumsum(toss == "W")
)

data <- data |>
    expand_grid(p_water = seq(0, 1, length.out = 50)) |>
    group_by(p_water) |>
    mutate(lagged_trial = lag(trial, n = 1),
           lagged_n_successes = lag(n_successes, n = 1)) |>
    ungroup() |>
    mutate(
        prior = ifelse(trial == 1, .5, dbinom(lagged_n_successes, lagged_trial, p_water)),
        likelihood = dbinom(n_successes, trial, p_water),
        strip = str_c("n = ", trial)
    ) |>
    group_by(trial) |>
    mutate(prior = prior / sum(prior),
           likelihood = likelihood / sum(likelihood)) |>
    ungroup()
```


```{r}
data |>
    ggplot(aes(x = p_water)) +
    geom_line(aes(y = prior), linetype = "dashed") +
    geom_line(aes(y = likelihood)) +
    facet_wrap(~strip, scales = "free_y")
```

## 2.6 Practice

### 2E1
Which of the expressions below correspond to the statement: *the probability of rain on Monday?*

1. Pr(rain)
2. Pr(rain|Monday)
3. Pr(Monday|rain)
4. Pr(rain, Monday) / Pr(Monday)

Expression 2 is correct.

### 2E2
Which the following statements corresponds to the expression: Pr(Monday|rain)?

1. The probability of rain on Monday.
2. The probability of rain, given that it is Monday.
3. The probability of Monday, given that it is raining.
4. The probability that it is Monday and that it is raining.

Expression 3 is correct.

### 2E3
Which of the expressions below correspond to the statement: *the probability that it is Monday, given that it is raining?*

1. Pr(Monday|rain)
2. Pr(rain|Monday)
3. Pr(rain|Monday) Pr(Monday)
4. Pr(rain|Monday) Pr(Monday)/Pr(rain)
5. Pr(Monday|rain) Pr(rain)/Pr(Monday)

Expression 4 is correct.

### 2E4

### 2M1
Recall the globe tossing model from the chapter. Compute and plot the grid approximate distribution for each of the following sets of observations. In each case, assume a uniform prior for $p$.

1. W, W, W
2. W, W, W, L
3. L, W, W, L, W, W, W


```{r}
grid_size <- 50
data <- list(
    obs_1 = c(1, 1, 1),
    obs_2 = c(1, 1, 1, 0),
    obs_3 = c(0, 1, 1, 0, 1, 1, 1)
)

tosses <- map(
    .x = data,
    .f = \(obs) {
        w <- sum(obs)
        n <- length(obs)
        grid_approximation(uniform_prior, grid_size, w, n)
    }
) |>
    bind_rows(.id = "obs")

tosses |>
    ggplot(aes(x = p_grid, y = posterior)) +
    geom_line() +
    facet_wrap(~obs, scales = "free_y")
```

### 2M2
Now assume a prior for $p$ that is equal to zero when $p < 0.5$ and is a positive constant when $p \geq 0.5$. Again, compute and plot the grid approximate distribution for each of the following sets of observations in the problem just above.

```{r}
tosses <- map(
    .x = data,
    .f = \(obs) {
        w <- sum(obs)
        n <- length(obs)
        grid_approximation(truncated_prior, grid_size, w, n)
    }
) |>
    bind_rows(.id = "obs")

tosses |>
    ggplot(aes(x = p_grid, y = posterior)) +
    geom_line() +
    facet_wrap(~obs, scales = "free_y")
```

### 2M3
Suppose there are two globes, one for Earth an one for Mars. The Earth glove is 70% covered in water. The Mars glove is 100% land. Further suppose that one of these gloves - you don't know which - was tosses in the air and produced a "land" observation. Assume that each glove was equally likely to be tossed. Show that the posterior probability that the globe was the Earth, conditional on seeing "land" (Pr(Earth|land)) is 0.23.